---
title: "Confidence Intervals"
categories: [statistics, procedure, foundations]
tier-focus: procedure
related:
  - ../../wiki/concepts/confidence-intervals.md
  - ../../wiki/concepts/standard-error.md
  - ../../wiki/concepts/distributions.md
---

# Confidence Intervals

## Learning Goals

1. State the correct interpretation of a confidence interval
2. Choose Z vs t based on what is known
3. Construct CI on μ (Z and t), σ², and p
4. Determine sample size to achieve a target margin of error

## In Plain English

A confidence interval gives a **range of plausible values** for an unknown population parameter. "95% confident" means: if we repeated this process many times, 95% of the intervals we build would contain the true parameter.

> **Teacher's phrasing:** "We are [100(1−α)]% confident that [parameter in context] is between [lower] and [upper]."

**It does NOT mean there is a 95% chance μ is in this particular interval.**

## General Structure

$$\text{Estimate} \pm \text{Critical value} \times \text{Standard Error}$$

## When to Use Z vs t

| Condition | Use |
|-----------|-----|
| σ known | Z |
| σ unknown, normal population | t (df = n−1) |
| σ unknown, n ≥ 40 | Z (CLT approximation) |

## CI on Mean μ

```{r}
#| label: ci-on-mean

# One-sample t CI (σ unknown)
set.seed(7)
x <- rnorm(20, mean = 100, sd = 15)

n    <- length(x)
xbar <- mean(x)
s    <- sd(x)
se   <- s / sqrt(n)
t_crit <- qt(0.975, df = n - 1)

lower <- xbar - t_crit * se
upper <- xbar + t_crit * se

cat(sprintf("x̄ = %.3f,  s = %.3f,  n = %d\n", xbar, s, n))
cat(sprintf("t* = %.3f\n", t_crit))
cat(sprintf("95%% CI: (%.3f, %.3f)\n", lower, upper))

# Compare with t.test()
t.test(x, conf.level = 0.95)$conf.int
```

## CI on Variance σ²

$$\frac{(n-1)s^2}{\chi^2_{\alpha/2,\,n-1}} \leq \sigma^2 \leq \frac{(n-1)s^2}{\chi^2_{1-\alpha/2,\,n-1}}$$

```{r}
#| label: ci-on-variance

n <- 20; s2 <- var(x)
chi_lo <- qchisq(0.025, df = n - 1)  # lower χ² critical value
chi_hi <- qchisq(0.975, df = n - 1)  # upper χ² critical value

cat(sprintf("95%% CI on σ²: (%.3f, %.3f)\n",
            (n-1)*s2/chi_hi, (n-1)*s2/chi_lo))
cat(sprintf("95%% CI on σ:  (%.3f, %.3f)\n",
            sqrt((n-1)*s2/chi_hi), sqrt((n-1)*s2/chi_lo)))
```

## Sample Size for Target Margin of Error

$$n = \left(\frac{z_{\alpha/2}\,\sigma}{E}\right)^2$$

```{r}
#| label: sample-size

# Required n to estimate μ within E = 5, with 95% confidence, σ ≈ 15
E     <- 5
sigma <- 15
z     <- qnorm(0.975)

n_req <- ceiling((z * sigma / E)^2)
cat("Required n:", n_req)
```

## Common Mistakes

- Using Z when σ is unknown and n is small
- Forgetting to take square root for CI on σ (vs σ²)
- Using α instead of α/2 for two-sided intervals
- Misinterpreting the CI as P(μ ∈ CI) = 0.95

## Try It Yourself

1. Change `conf.level` to 0.99. How does the CI width change?
2. For the χ² CI: why are the bounds **not** symmetric around s²?
3. Double the sample size n. By what factor does the CI width shrink?

## Related Wiki Pages

- [Confidence Intervals (formulas)](../../wiki/concepts/confidence-intervals.md)
- [Standard Error](../../wiki/concepts/standard-error.md)
- [Hypothesis Testing Overview](../../wiki/concepts/hypothesis-testing-overview.md)
