---
title: "Sampling Distributions and the CLT"
categories: [statistics, construct, foundations]
tier-focus: construct
related:
  - ../../wiki/concepts/distributions.md
  - ../../wiki/concepts/probability.md
  - ../../wiki/concepts/standard-error.md
---

# Sampling Distributions and the CLT

## Learning Goals

1. Explain the difference between a population distribution and a sampling distribution
2. State the Central Limit Theorem and identify when it applies
3. Recognize the normal, t, χ², and F distributions and when each is used
4. Compute normal and t probabilities in R

## In Plain English

A **sampling distribution** is the distribution of a statistic (like x̄) across all possible samples of size n. You never see it directly — you only see one sample — but understanding it is the foundation of all inference.

The **Central Limit Theorem (CLT)** says: for large enough n, x̄ is approximately normal regardless of the population distribution.

$$\bar{X} \sim N\!\left(\mu,\; \frac{\sigma^2}{n}\right) \quad \text{(approximately, for large n)}$$

This is why we can make probability statements about sample means.

## The Four Key Distributions

| Distribution | Used for | df |
|-------------|---------|-----|
| Normal Z | Mean when σ known, or n ≥ 40 | — |
| Student's t | Mean when σ unknown, n small | n−1 |
| Chi-square χ² | Variance, goodness-of-fit | varies |
| F | Ratio of variances, ANOVA, regression | (df₁, df₂) |

## R Example: CLT Simulation

```{r}
#| label: clt-simulation
#| fig-cap: "Sampling distribution of x̄ from a skewed population, n=30"

set.seed(123)

# Skewed population (exponential)
pop <- rexp(100000, rate = 1/10)  # mean = 10

# Draw 5000 samples of size n=30, compute each x̄
n_sim  <- 5000
n_samp <- 30
xbars  <- replicate(n_sim, mean(sample(pop, n_samp)))

# Plot
hist(xbars, breaks = 50, col = "steelblue", border = "white",
     main = paste("Sampling dist of x̄ (n =", n_samp, ")"),
     xlab = "Sample mean", probability = TRUE)
curve(dnorm(x, mean = mean(pop), sd = sd(pop)/sqrt(n_samp)),
      add = TRUE, col = "firebrick", lwd = 2)
```

## Normal and t Probability in R

```{r}
#| label: probability-r

# P(Z < 1.96)
pnorm(1.96)

# P(Z > 1.96)  [upper tail]
pnorm(1.96, lower.tail = FALSE)

# t critical value: df = 15, α/2 = 0.025
qt(0.975, df = 15)

# P(T₁₅ < 2.131)
pt(2.131, df = 15)
```

## Try It Yourself

1. Change `n_samp` from 30 to 5. Does the sampling distribution still look normal?
2. Change `rexp` to `runif(100000, 0, 20)` (uniform population). Does CLT still hold?
3. What sample size is "large enough" for the CLT to kick in for an exponential population?

## Related Wiki Pages

- [Distributions](../../wiki/concepts/distributions.md)
- [Probability](../../wiki/concepts/probability.md)
- [Standard Error](../../wiki/concepts/standard-error.md)
