---
title: "One-Way ANOVA"
categories: [statistics, procedure, inference]
tier-focus: procedure
related:
  - ../../wiki/concepts/crd-one-way-anova.md
  - ../../wiki/concepts/ss-decomposition.md
  - ../../wiki/r-code/anova-r.md
---

# One-Way ANOVA

## Learning Goals

1. State when one-way ANOVA is appropriate (a ≥ 3 groups, single factor)
2. Construct and interpret the ANOVA table (SS, df, MS, F₀)
3. Apply Fisher's LSD for pairwise comparisons after a significant F-test
4. Run a CRD ANOVA in R and extract all required values

## In Plain English

One-way ANOVA tests whether **a ≥ 3 group means are all equal**. It decomposes total variation (SS_T) into variation **between** groups (SS_Treatments) and variation **within** groups (SS_Error).

$$F_0 = \frac{MS_{Treatments}}{MS_E}$$

A large F₀ means between-group variation is large relative to within-group noise — evidence against H₀: μ₁ = μ₂ = ... = μₐ.

## ANOVA Table Structure

| Source | SS | df | MS | F₀ |
|--------|----|----|----|----|
| Treatments | SS_T | a−1 | SS_T/(a−1) | MS_Trt/MS_E |
| Error | SS_E | N−a | SS_E/(N−a) | |
| Total | SS_T | N−1 | | |

$\hat{\sigma} = \sqrt{MS_E}$

## Fisher's LSD (Post-Hoc)

After rejecting H₀, test all pairwise differences:

$$LSD = t_{\alpha/2,\, N-a} \sqrt{MS_E\!\left(\frac{1}{n_i}+\frac{1}{n_j}\right)}$$

Reject H₀: μᵢ = μⱼ if |ȳᵢ − ȳⱼ| > LSD.

## R Example

```{r}
#| label: one-way-anova

pacman::p_load(pacman, tidyverse)

# Tensile strength of concrete with 4 mixing methods (balanced, n=5 each)
set.seed(9)
df <- data.frame(
  strength = c(rnorm(5, 3000, 150), rnorm(5, 3200, 150),
               rnorm(5, 3100, 150), rnorm(5, 3350, 150)),
  method   = factor(rep(c("A","B","C","D"), each = 5))
)

# Fit one-way ANOVA (CRD)
fit <- aov(strength ~ method, data = df)
summary(fit)
```

```{r}
#| label: anova-extraction

# Extract key values
aov_tbl <- summary(fit)[[1]]
rownames(aov_tbl) <- trimws(rownames(aov_tbl))

MSE  <- aov_tbl["Residuals", "Mean Sq"]
df_E <- aov_tbl["Residuals", "Df"]
N    <- nrow(df)
a    <- nlevels(df$method)
n    <- N / a

cat("MSE =", round(MSE, 2), "\n")
cat("σ̂  =", round(sqrt(MSE), 2), "\n")

# Fisher LSD (α = 0.05)
LSD <- qt(0.975, df = df_E) * sqrt(MSE * (1/n + 1/n))
cat("LSD =", round(LSD, 2), "\n")

# Group means
group_means <- tapply(df$strength, df$method, mean)
print(round(group_means, 2))
```

## Try It Yourself

1. Which pairs of methods have significantly different means (|ȳᵢ − ȳⱼ| > LSD)?
2. Increase within-group SD from 150 to 300. Does the F-test still reject H₀?
3. What does MS_E estimate? What does √MS_E represent?

## Related Wiki Pages

- [CRD and One-Way ANOVA](../../wiki/concepts/crd-one-way-anova.md)
- [SS Decomposition](../../wiki/concepts/ss-decomposition.md)
- [R: ANOVA](../../wiki/r-code/anova-r.md)
- [ANOVA Table Examples](../../wiki/examples/anova-table-examples.md)
