---
title: "Model Diagnostics and Adequacy"
categories: [statistics, script, regression]
tier-focus: script
related:
  - ../../wiki/concepts/model-adequacy.md
  - ../../wiki/examples/model-adequacy-examples.md
  - ../../wiki/r-code/regression-r.md
---

# Model Diagnostics and Adequacy

## Learning Goals

1. Produce the four standard residual diagnostic plots
2. Diagnose violations of linearity, constant variance, and normality
3. Identify outliers (large residuals), leverage points, and influential observations
4. State conclusions using the teacher's expected diagnostic phrasing

## The Four Assumptions to Check

| Assumption | What to look for | Plot |
|-----------|-----------------|------|
| Linearity | No curve in residuals vs fitted | Residuals vs ŷ |
| Constant variance | No fan shape | Residuals vs ŷ |
| Normality | Points follow the line | Normal Q-Q plot |
| Independence | No pattern vs. time/order | Residuals vs run order |

## Teacher's Phrasing

> "The residuals show [random scatter / a curved pattern / a fan shape], suggesting [no violation / a violation of linearity / a violation of constant variance] of the [linearity / constant variance] assumption."

## R Example: Full Diagnostic Workflow

```{r}
#| label: diagnostics-setup

pacman::p_load(pacman, tidyverse)

# Use SLR rocket motor model from simple-linear-regression lesson
age      <- c(15.5, 23.75, 8, 17, 5.5, 19, 24, 2.5, 7.5, 11, 13, 3.75)
strength <- c(2158.7, 1678.15, 2316, 2061.3, 2207.5, 1708.3,
              1784.7, 2575, 2357.9, 2256.7, 2165.2, 2399.55)
df  <- tibble(age, strength)
fit <- lm(strength ~ age, data = df)

# Add residuals and diagnostics to data frame
df <- df |>
  mutate(resid  = residuals(fit),
         fitted = fitted(fit),
         rstud  = rstudent(fit),
         h      = hatvalues(fit))
```

```{r}
#| label: residuals-vs-fitted
#| fig-cap: "Residuals vs Fitted — check linearity and constant variance"

ggplot(df, aes(x = fitted, y = resid)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick") +
  labs(x = "Fitted Values", y = "Residuals",
       title = "Residuals vs Fitted") +
  theme_bw()
```

```{r}
#| label: qq-plot
#| fig-cap: "Normal Q-Q — check normality of residuals"

ggplot(df, aes(sample = resid)) +
  stat_qq(size = 2.5, color = "steelblue") +
  stat_qq_line(color = "firebrick") +
  labs(title = "Normal Q-Q Plot of Residuals") +
  theme_bw()
```

```{r}
#| label: shapiro-wilk

# Shapiro-Wilk normality test
sw <- shapiro.test(df$resid)
cat("Shapiro-Wilk: W =", round(sw$statistic, 4),
    ", p-value =", round(sw$p.value, 4), "\n")

if (sw$p.value > 0.05) {
  cat("Normality assumption is reasonable (fail to reject at α=0.05).\n")
} else {
  cat("Normality assumption may be violated (reject at α=0.05).\n")
}
```

```{r}
#| label: influence

# Cook's distance — identify influential observations
cooks <- cooks.distance(fit)
cat("Top Cook's distances:\n")
sort(cooks, decreasing = TRUE) |> head(5) |> round(4) |> print()

threshold <- 4 / nrow(df)
cat("\nInfluential (Cook's D > 4/n =", round(threshold, 4), "):\n")
which(cooks > threshold)
```

## Try It Yourself

1. Does the Q-Q plot suggest any departure from normality?
2. Is there any evidence of non-constant variance in the residuals vs fitted plot?
3. Are any observations influential based on Cook's distance?

## Related Wiki Pages

- [Model Adequacy](../../wiki/concepts/model-adequacy.md)
- [Model Adequacy Examples](../../wiki/examples/model-adequacy-examples.md)
- [R: Regression Workflow](../../wiki/r-code/regression-r.md)
