---
title: "p, c, and u Charts"
categories: [statistics, script, quality]
tier-focus: script
related:
  - ../../wiki/concepts/control-charts.md
  - ../../wiki/r-code/control-charts-r.md
---

# p, c, and u Charts

## Learning Goals

1. Choose among p, c, and u charts based on data type and sample size
2. Build each chart in R using the qcc package
3. Handle variable sample sizes in p and u charts
4. Interpret out-of-control signals on attribute charts

## Chart Selection

| Chart | Data | Sample size |
|-------|------|------------|
| p | Fraction defective (proportion) | Variable or fixed |
| c | Count of defects (per inspection unit) | Fixed |
| u | Defects per unit | Variable |

## p-Chart (Fraction Defective)

$$\bar{p} = \frac{\sum D_i}{\sum n_i}, \quad UCL_p = \bar{p} + 3\sqrt{\frac{\bar{p}(1-\bar{p})}{n_i}}, \quad LCL_p = \max(0, \bar{p} - \cdots)$$

With variable n, limits vary by subgroup.

## R Example: p-Chart

```{r}
#| label: p-chart

pacman::p_load(pacman, qcc)

data("orangejuice")  # D = defectives, size = sample size
head(orangejuice)

p_chart <- qcc(data = orangejuice$D, type = "p",
               sizes = orangejuice$size,
               title = "p-Chart — Orange Juice Cans")

cat("\np-bar:", round(p_chart$center, 4), "\n")
cat("Out-of-control points:", p_chart$violations$beyond.limits, "\n")
```

## R Example: c-Chart and u-Chart

```{r}
#| label: c-u-charts

data("circuit")  # x = defects, size = inspection area
head(circuit)

# c-chart (fixed sample size)
c_chart <- qcc(data = circuit$x[circuit$size == 1],
               type = "c",
               title = "c-Chart — Circuit Boards")

# u-chart (variable area)
u_chart <- qcc(data = circuit$x, type = "u",
               sizes = circuit$size,
               title = "u-Chart — Circuit Boards")
```

## Try It Yourself

1. For the p-chart: which subgroups are out of control?
2. For the u-chart: why do control limits vary from subgroup to subgroup?
3. Remove the out-of-control points from the p-chart and rebuild with revised limits.

## Related Wiki Pages

- [Control Charts](../../wiki/concepts/control-charts.md)
- [R: Control Charts](../../wiki/r-code/control-charts-r.md)
- [Control Chart Examples](../../wiki/examples/control-chart-examples.md)
