---
title: "X̄/R and X̄/S Charts"
categories: [statistics, script, quality]
tier-focus: script
related:
  - ../../wiki/concepts/control-charts.md
  - ../../wiki/r-code/control-charts-r.md
  - ../../wiki/examples/control-chart-examples.md
---

# X̄/R and X̄/S Charts

## Learning Goals

1. Build X̄/R charts in R using the qcc package (R chart first, always)
2. Identify out-of-control points and revise limits
3. Compute Cp and Cpk and interpret process capability
4. State conclusions using the teacher's expected phrasing

## Fixed Sequence (Script)

1. Build **R chart first** — estimate σ̂ from R̄/d₂
2. Build **X̄ chart** using σ̂ from the R chart
3. Identify any out-of-control points
4. Remove special-cause points and **revise limits**
5. Compute **Cp and Cpk** against specification limits
6. State process capability conclusion

## R Example: Variables Control Chart

```{r}
#| label: xbar-r

pacman::p_load(pacman, qcc, tidyverse)

# Piston ring diameter data (n=5 per subgroup, 25 subgroups)
# Using built-in pistonrings dataset
data("pistonrings")

diameter <- qcc.groups(pistonrings$diameter, pistonrings$sample)

# Step 1: R chart
r_chart <- qcc(diameter, type = "R", nsigmas = 3,
               title = "R Chart — Piston Ring Diameter")
```

```{r}
#| label: xbar-chart

# Step 2: X̄ chart (σ̂ from R chart)
xbar_chart <- qcc(diameter, type = "xbar", nsigmas = 3,
                  std.dev = sd.R(diameter),
                  title = "X̄ Chart — Piston Ring Diameter")

cat("\nCenter (x̄̄):", round(xbar_chart$center, 4), "\n")
cat("UCL:", round(xbar_chart$limits[2], 4), "\n")
cat("LCL:", round(xbar_chart$limits[1], 4), "\n")
cat("σ̂ :", round(xbar_chart$std.dev, 6), "\n")
```

```{r}
#| label: process-capability

# Process capability (LSL = 73.95, USL = 74.05 mm)
pc <- process.capability(xbar_chart, spec.limits = c(73.95, 74.05))
print(pc, digits = 6)
```

## Teacher's Phrasing

| Finding | What to say |
|---------|------------|
| All points in limits | "The process is in statistical control." |
| Points outside limits | "Point(s) [#] fall outside the control limits, indicating the process is out of control." |
| Cp ≥ 1.33, Cpk ≥ 1.33 | "The process is capable of meeting specifications." |
| Cp ≥ 1.33, Cpk < 1.33 | "The process spread is adequate but the process is off-center." |
| Cp < 1.33 | "The process is not capable of meeting specifications." |

## Try It Yourself

1. Which points (if any) are flagged as out of control on the X̄ chart?
2. What is Cpk? Is the process capable?
3. Remove any out-of-control subgroups and rebuild the charts with revised limits.

## Related Wiki Pages

- [Control Charts](../../wiki/concepts/control-charts.md)
- [Process Capability](../../wiki/concepts/process-capability.md)
- [R: Control Charts](../../wiki/r-code/control-charts-r.md)
