---
title: "CRD and One-Way ANOVA"
categories: [statistics, procedure, doe]
tier-focus: procedure
related:
  - ../../wiki/concepts/crd-one-way-anova.md
  - ../../wiki/concepts/rcbd-blocking.md
  - ../../wiki/r-code/anova-r.md
---

# Completely Randomized Design (CRD) and One-Way ANOVA

## Learning Goals

1. Identify when a CRD is appropriate and when blocking is needed
2. Set up the one-way ANOVA model and construct the ANOVA table
3. Apply Fisher's LSD for pairwise comparisons
4. Run both balanced and unbalanced CRD in R and extract key values

## In Plain English

A CRD randomly assigns experimental units to treatment levels. Use CRD when experimental units are **homogeneous**. If there is a known nuisance variable (time, batch, operator), use RCBD instead — blocking removes that variation from SS_Error.

## ANOVA Table (Balanced CRD)

| Source | SS | df | MS | F₀ |
|--------|----|----|----|----|
| Treatments | SS_Trt | a−1 | MS_Trt | MS_Trt/MS_E |
| Error | SS_E | N−a | MS_E | |
| Total | SS_T | N−1 | | |

$$\hat{\sigma} = \sqrt{MS_E}, \quad \text{Fisher LSD} = t_{\alpha/2,\,N-a}\sqrt{MS_E\left(\tfrac{1}{n_i}+\tfrac{1}{n_j}\right)}$$

## R Example

```{r}
#| label: crd-r

pacman::p_load(pacman, rio, tidyverse)

# Coating thickness for 4 application methods (5 replicates each)
set.seed(15)
df <- data.frame(
  thickness = c(rnorm(5, 125, 8), rnorm(5, 132, 8),
                rnorm(5, 119, 8), rnorm(5, 138, 8)),
  method    = factor(rep(c("Spray","Dip","Brush","Roll"), each = 5))
)

# Fit ANOVA
fit <- aov(thickness ~ method, data = df)

# ANOVA table
aov_tbl <- summary(fit)[[1]]
rownames(aov_tbl) <- trimws(rownames(aov_tbl))
print(aov_tbl)
```

```{r}
#| label: crd-lsd

# Extract values
MSE  <- aov_tbl["Residuals", "Mean Sq"]
df_E <- aov_tbl["Residuals", "Df"]
N    <- nrow(df); a <- nlevels(df$method); n <- N / a

cat("MSE =", round(MSE, 3), "\n")
cat("σ̂  =", round(sqrt(MSE), 3), "\n")
cat("F₀  =", round(aov_tbl["method", "F value"], 3), "\n")
cat("p   =", round(aov_tbl["method", "Pr(>F)"], 4), "\n\n")

# Fisher LSD
LSD <- qt(0.975, df = df_E) * sqrt(MSE * 2/n)
cat("LSD =", round(LSD, 3), "\n\n")

# Group means and pairwise differences
means <- tapply(df$thickness, df$method, mean)
cat("Group means:\n"); print(round(means, 2))
```

## RCBD: When to Block

If you suspect a nuisance factor (e.g., different operators, batches), use RCBD:

```{r}
#| label: rcbd-stub

# RCBD model: response ~ treatment + block
# fit_rcbd <- aov(y ~ treatment + block, data = df_rcbd)
# Key: df_Error = (a-1)(b-1) instead of N-a
cat("RCBD df_Error = (a-1)(b-1)\n")
cat("CRD df_Error  = N-a\n")
cat("Blocking removes SS_Block from SS_Error → smaller MS_E → larger F₀\n")
```

## Try It Yourself

1. Is the F-test significant at α = 0.05? Which methods differ significantly by LSD?
2. What does MS_E estimate in this context?
3. If you suspected that different operators applied each method, what design would you switch to?

## Related Wiki Pages

- [CRD and One-Way ANOVA](../../wiki/concepts/crd-one-way-anova.md)
- [RCBD Blocking](../../wiki/concepts/rcbd-blocking.md)
- [R: ANOVA](../../wiki/r-code/anova-r.md)
- [CRD Examples](../../wiki/examples/crd-examples.md)
