---
title: "Factorial ANOVA"
categories: [statistics, procedure, doe]
tier-focus: procedure
related:
  - ../../wiki/concepts/factorial-anova.md
  - ../../wiki/r-code/factorial-anova-r.md
  - ../../wiki/examples/factorial-examples.md
---

# Factorial ANOVA

## Learning Goals

1. Distinguish main effects from interaction effects
2. Construct the ANOVA table for a two-factor factorial design
3. Apply the 3× SD rule for constant variance check
4. Interpret an interaction plot and apply the marginality principle

## In Plain English

A factorial design tests multiple factors simultaneously and — critically — can detect **interactions**: situations where the effect of one factor depends on the level of another.

**Interaction present:** the optimal level of A depends on which level of B you're at.
**Interaction absent:** the optimal level of A is the same regardless of B.

## ANOVA Table (Two-Factor, Balanced)

| Source | df | MS | F₀ |
|--------|----|----|-------|
| A | a−1 | MS_A | MS_A/MS_E |
| B | b−1 | MS_B | MS_B/MS_E |
| AB (interaction) | (a−1)(b−1) | MS_AB | MS_AB/MS_E |
| Error | ab(n−1) | MS_E | |
| Total | abn−1 | | |

## Critical Rules

- **trimws() is mandatory** after `summary(fit)[[1]]` — R pads row names with trailing spaces
- **Hierarchical principle** — if AB is in the model, both A and B stay even if one is non-significant
- **3× SD rule** — if one cell's within-cell SD is ≈3× larger than others, constant variance is not reasonable

## R Example: Two-Factor Factorial

```{r}
#| label: factorial-fit

pacman::p_load(pacman, agricolae, tidyverse)

# Viscosity experiment: 2 pH levels × 2 catalyst types, n=4 per cell
set.seed(33)
df <- expand.grid(pH = factor(c("Low","High")),
                  catalyst = factor(c("A","B")),
                  rep = 1:4) |>
  mutate(viscosity = c(rnorm(4, 82, 3), rnorm(4, 91, 3),
                       rnorm(4, 79, 3), rnorm(4, 76, 3))) |>
  select(-rep)

fit <- aov(viscosity ~ pH * catalyst, data = df)

aov_tbl <- summary(fit)[[1]]
rownames(aov_tbl) <- trimws(rownames(aov_tbl))
print(aov_tbl)
```

```{r}
#| label: interaction-plot
#| fig-cap: "Interaction plot: if lines are parallel, no interaction"

df |>
  group_by(pH, catalyst) |>
  summarise(mean_visc = mean(viscosity), .groups = "drop") |>
  ggplot(aes(x = pH, y = mean_visc, color = catalyst, group = catalyst)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  labs(title = "Interaction Plot: Viscosity",
       y = "Mean Viscosity", x = "pH") +
  theme_bw()
```

```{r}
#| label: constant-variance-check

# 3× SD rule: within-cell SDs
df |>
  group_by(pH, catalyst) |>
  summarise(sd = sd(viscosity), .groups = "drop") |>
  print()

# LSD on marginal means of significant factor
MSE  <- aov_tbl["Residuals", "Mean Sq"]
df_E <- aov_tbl["Residuals", "Df"]
a    <- nlevels(df$pH); b <- nlevels(df$catalyst); n <- 4
n_i  <- b * n   # observations per level of pH (marginal mean)

LSD_pH <- qt(0.975, df_E) * sqrt(MSE * 2 / n_i)
cat("LSD for pH marginal means:", round(LSD_pH, 3), "\n")
```

## Try It Yourself

1. Is the interaction significant? How do you interpret the interaction plot?
2. Apply the 3× SD rule — is constant variance reasonable?
3. If the interaction is not significant, which factor should drive recommendations?

## Related Wiki Pages

- [Factorial ANOVA](../../wiki/concepts/factorial-anova.md)
- [R: Factorial ANOVA](../../wiki/r-code/factorial-anova-r.md)
- [Factorial Examples](../../wiki/examples/factorial-examples.md)
