---
title: "Drill: Build a Control Chart"
categories: [r, drill, coding-dojo]
tier-focus: script
---

# Drill: Build a Control Chart

## Drill Purpose

Build fluency with the qcc control chart workflow: R chart → X̄ chart → identify out-of-control → revise limits → process capability.

## Trigger

Use this drill when you have subgroup measurement data and need to determine whether a process is in statistical control and whether it meets specification limits.

## Fixed Sequence

1. Load data (subgroup matrix: rows = subgroups, cols = observations)
2. Build **R chart first** — estimates σ̂
3. Build **X̄ chart** using σ̂ from R chart
4. Identify out-of-control subgroups (`$violations`)
5. Remove and **revise limits**
6. Compute **Cp and Cpk** against spec limits
7. State conclusion in teacher's phrasing

## Setup

```{r}
#| label: setup

pacman::p_load(pacman, qcc)
```

## Worked Example: Piston Rings

```{r}
#| label: r-chart
#| fig-cap: "R Chart — check for out-of-control range"

data("pistonrings")
diameter <- qcc.groups(pistonrings$diameter, pistonrings$sample)

r_chart <- qcc(diameter, type = "R", nsigmas = 3,
               title = "R Chart — Piston Ring Diameter")

cat("R̄  (center):", round(r_chart$center, 6), "\n")
cat("UCL_R:", round(r_chart$limits[2], 6), "\n")
cat("R chart violations:", r_chart$violations$beyond.limits, "\n")
```

```{r}
#| label: xbar-chart
#| fig-cap: "X̄ Chart — check for out-of-control means"

xbar_chart <- qcc(diameter, type = "xbar", nsigmas = 3,
                  std.dev = sd.R(diameter),
                  title = "X̄ Chart — Piston Ring Diameter")

cat("x̄̄  (center):", round(xbar_chart$center, 6), "\n")
cat("UCL:", round(xbar_chart$limits[2], 6), "\n")
cat("LCL:", round(xbar_chart$limits[1], 6), "\n")
cat("σ̂ :", round(xbar_chart$std.dev, 6), "\n")
cat("X̄ chart violations:", xbar_chart$violations$beyond.limits, "\n")
```

```{r}
#| label: capability
#| fig-cap: "Process capability"

# Spec limits: LSL = 73.95, USL = 74.05
pc <- process.capability(xbar_chart, spec.limits = c(73.95, 74.05))
print(pc, digits = 6)
```

## Practice Variation: p-Chart

```{r}
#| label: p-chart
#| fig-cap: "p-Chart — fraction defective"

data("orangejuice")

p_chart <- qcc(data  = orangejuice$D,
               type  = "p",
               sizes = orangejuice$size,
               title = "p-Chart — Orange Juice")

cat("\np̄:", round(p_chart$center, 4), "\n")
cat("Out-of-control:", p_chart$violations$beyond.limits, "\n")
```

## Reflection Questions

1. For the piston ring X̄ chart: is the process in statistical control? In control but not capable?
2. Is Cpk ≥ 1.33? What does that mean for the process?
3. For the p-chart: which lots should be investigated for special causes?

## Related Wiki Pages

- [R: Control Charts](../../wiki/r-code/control-charts-r.md)
- [Control Charts](../../wiki/concepts/control-charts.md)
- [Process Capability](../../wiki/concepts/process-capability.md)
