---
title: "Drill: Factorial ANOVA in R"
categories: [r, drill, coding-dojo]
tier-focus: script
---

# Drill: Factorial ANOVA in R

## Drill Purpose

Build fluency with the two-factor and 2^k ANOVA workflow in R: import wide data → reshape long → fit model → trimws() → extract values → interaction plot → LSD.

## Trigger

Use this drill when you have a factorial experiment with 2+ factors and need to test main effects, interactions, and do pairwise comparisons.

## Fixed Sequence

1. Import (wide format: one column per replicate)
2. `pivot_longer()` — reshape to long format
3. `aov()` — fit factorial model
4. `summary()[[1]]` + **`trimws()`** — extract ANOVA table
5. Check interaction significance
6. If interaction non-significant: `LSD.test()` on significant main effect marginal means
7. Interaction plot — interpret parallel vs crossing lines

## Setup

```{r}
#| label: setup

pacman::p_load(pacman, agricolae, tidyverse)
```

## Worked Example: Two-Factor Warping Data

```{r}
#| label: import-reshape

# Warping of copper plates: factor A = temperature (4 levels), B = Cu content (3 levels)
# Entered in wide format (R1, R2 = replicates)
df_wide <- tribble(
  ~temp, ~cu,  ~R1,  ~R2,
  "50",  "40",  17,   20,
  "50",  "60",  12,    9,
  "50",  "80",  16,   12,
  "50", "100",  21,   17,
  "75",  "40",  12,   13,
  "75",  "60",  18,   21,
  "75",  "80",  10,   12,
  "75", "100",  24,   21,
  "100", "40",  14,   15,
  "100", "60",  18,   20,
  "100", "80",  14,   11,
  "100","100",  23,   22,
  "125", "40",  19,   17,
  "125", "60",  23,   20,
  "125", "80",  19,   16,
  "125","100",  24,   28
)

# Reshape to long
df <- df_wide |>
  pivot_longer(cols = c(R1, R2), names_to = "rep", values_to = "warping") |>
  mutate(temp = factor(temp), cu = factor(cu))

head(df)
```

```{r}
#| label: fit-aov

fit <- aov(warping ~ temp * cu, data = df)

aov_tbl <- summary(fit)[[1]]
rownames(aov_tbl) <- trimws(rownames(aov_tbl))  # MANDATORY
print(aov_tbl)
```

```{r}
#| label: extract-values

MSE  <- aov_tbl["Residuals", "Mean Sq"]
df_E <- aov_tbl["Residuals", "Df"]
N    <- nrow(df); a <- nlevels(df$temp); b <- nlevels(df$cu); n <- 2

cat("MSE =", round(MSE, 4), "\n")
cat("σ̂  =", round(sqrt(MSE), 4), "\n")

# Is interaction significant?
p_int <- aov_tbl["temp:cu", "Pr(>F)"]
cat("Interaction p-value:", round(p_int, 4), "\n")
```

```{r}
#| label: interaction-plot
#| fig-cap: "Interaction plot: parallel lines → no interaction"

df |>
  group_by(temp, cu) |>
  summarise(mean_w = mean(warping), .groups = "drop") |>
  ggplot(aes(x = cu, y = mean_w, color = temp, group = temp)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  labs(title = "Warping: Interaction Plot",
       x = "Cu Content (%)", y = "Mean Warping", color = "Temp (°C)") +
  theme_bw()
```

```{r}
#| label: lsd-test

# If interaction not significant, apply LSD on significant main effect
# LSD for Cu (b=4 levels, n_i = a*n observations per Cu level)
n_i_cu <- a * n
LSD_cu  <- qt(0.975, df_E) * sqrt(MSE * 2 / n_i_cu)
cat("\nLSD for Cu marginal means:", round(LSD_cu, 4), "\n")

# Marginal means for Cu
tapply(df$warping, df$cu, mean) |> round(3) |> print()
```

## Reflection Questions

1. Is the interaction significant at α = 0.05? What does this mean for your recommendations?
2. Apply the 3× SD rule to the within-cell standard deviations. Is constant variance reasonable?
3. For each significant main effect, which levels are significantly different by LSD?

## Related Wiki Pages

- [R: Factorial ANOVA](../../wiki/r-code/factorial-anova-r.md)
- [Factorial ANOVA](../../wiki/concepts/factorial-anova.md)
- [Factorial Examples](../../wiki/examples/factorial-examples.md)
