---
title: "Drill: Fit and Interpret a Regression Model"
categories: [r, drill, coding-dojo]
tier-focus: script
filters:
  - webr
webr:
  packages:
    - ggplot2
    - tibble
    - dplyr
---

# Drill: Fit and Interpret a Regression Model

## Drill Purpose

Build fluency with the complete regression workflow: fit → ANOVA → summary → confint → predict → diagnostics. This sequence appears in full on exams.

## Trigger

Use this drill when you see: a dataset with one continuous response and one or more predictors, and you need to fit a regression model, test coefficients, build intervals, and check assumptions.

## Fixed Sequence

1. Import and inspect data
2. `lm()` — fit the model
3. `anova()` — ANOVA table (F-test, SS decomposition)
4. `summary()` — coefficient table, R², σ̂
5. `confint()` — CI on βⱼ (watch column names!)
6. `predict()` — CI on mean response and PI on future observation
7. Residual diagnostics — check assumptions

## Setup

```{r}
#| label: setup

pacman::p_load(pacman, tidyverse)
```

## Worked Example: Wire Bond Data

```{r}
#| label: import

# Wire bond pull strength data (Montgomery Table 12.1)
df <- tribble(
  ~strength, ~length, ~height,
   9.95, 2,  50,  24.45, 8, 110,  31.75, 11, 120,  35.00, 10, 550,
  25.02, 8, 295,  16.86, 4, 200,  14.38,  2, 375,   9.60,  2,  52,
  24.35, 9, 100,  27.50, 8, 300,  17.08,  4, 412,  37.00, 11, 400,
  41.95,12, 500,  11.66, 2, 360,  21.65,  4, 205,  17.89,  4, 400,
  69.00,20, 600,  10.30, 1, 585,  34.93, 10, 540,  46.59, 15, 250
)

glimpse(df)
```

```{r}
#| label: fit

fit <- lm(strength ~ length + height, data = df)
```

```{r}
#| label: anova-table

anova(fit)

# Extract SSR, SSE, MSE manually
a   <- anova(fit)
SSR <- sum(a[1:(nrow(a)-1), "Sum Sq"])
SSE <- a["Residuals", "Sum Sq"]
MSE <- a["Residuals", "Mean Sq"]
cat("SSR =", round(SSR, 3), "  SSE =", round(SSE, 3),
    "  MSE =", round(MSE, 3), "\n")
```

```{r}
#| label: summary-table

summary(fit)

# Key scalars
cat("\nR²     =", round(summary(fit)$r.squared, 4), "\n")
cat("R²_adj =", round(summary(fit)$adj.r.squared, 4), "\n")
cat("σ̂     =", round(summary(fit)$sigma, 4), "\n")
```

```{r}
#| label: confint

# 95% CI on each coefficient
ci <- confint(fit, level = 0.95)
print(ci)

# Extraction by name (critical: use exact column name)
cat("\nβ̂₁ (length) CI lower:", round(ci["length", "2.5 %"], 4), "\n")
cat("β̂₁ (length) CI upper:", round(ci["length", "97.5 %"], 4), "\n")
```

```{r}
#| label: predict

# Predict at length = 10, height = 300
x0 <- data.frame(length = 10, height = 300)

ci_mean <- predict(fit, newdata = x0, interval = "confidence", level = 0.95)
pi_obs  <- predict(fit, newdata = x0, interval = "prediction",  level = 0.95)

cat("Point estimate:", round(ci_mean[,"fit"], 3), "\n")
cat("95% CI on mean: (", round(ci_mean[,"lwr"], 3), ",",
                        round(ci_mean[,"upr"], 3), ")\n")
cat("95% PI on obs:  (", round(pi_obs[,"lwr"], 3), ",",
                        round(pi_obs[,"upr"], 3), ")\n")
```

```{r}
#| label: diagnostics
#| fig-cap: "Residual diagnostics"
#| fig-subcap:
#|   - "Residuals vs Fitted"
#|   - "Normal Q-Q"
#| layout-ncol: 2

df <- df |>
  mutate(resid = residuals(fit), fitted_val = fitted(fit))

ggplot(df, aes(x = fitted_val, y = resid)) +
  geom_point(color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Fitted", y = "Residuals") + theme_bw()

ggplot(df, aes(sample = resid)) +
  stat_qq(color = "steelblue") + stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q") + theme_bw()
```

## Your Turn

Modify the prediction point or fit a simple model on your own data. Click **Run Code** to execute.

```{webr-r}
#| label: sandbox

library(tibble)

# Wire bond data (same as worked example)
df <- tribble(
  ~strength, ~length, ~height,
   9.95, 2,  50,  24.45, 8, 110,  31.75, 11, 120,  35.00, 10, 550,
  25.02, 8, 295,  16.86, 4, 200,  14.38,  2, 375,   9.60,  2,  52,
  24.35, 9, 100,  27.50, 8, 300,  17.08,  4, 412,  37.00, 11, 400,
  41.95,12, 500,  11.66, 2, 360,  21.65,  4, 205,  17.89,  4, 400,
  69.00,20, 600,  10.30, 1, 585,  34.93, 10, 540,  46.59, 15, 250
)

fit <- lm(strength ~ length + height, data = df)

# Change these and re-run:
x0 <- data.frame(length = 10, height = 300)

ci_mean <- predict(fit, newdata = x0, interval = "confidence", level = 0.95)
pi_obs  <- predict(fit, newdata = x0, interval = "prediction",  level = 0.95)

cat("Point estimate:", round(ci_mean[,"fit"], 3), "\n")
cat("95% CI on mean: (", round(ci_mean[,"lwr"], 3), ",", round(ci_mean[,"upr"], 3), ")\n")
cat("95% PI on obs:  (", round(pi_obs[,"lwr"], 3),  ",", round(pi_obs[,"upr"], 3),  ")\n")
```

## Reflection Questions

1. What does β̂₁ = [value] mean in words?
2. Is the 95% CI on β̂₁ entirely positive/negative? What does that tell you?
3. Why is the PI always wider than the CI at the same x₀?

## Related Wiki Pages

- [R: Regression Workflow](../../wiki/r-code/regression-r.md)
- [Regression — SLR](../../wiki/concepts/regression-slr.md)
- [Regression — MLR](../../wiki/concepts/regression-mlr.md)
- [Regression CI Examples](../../wiki/examples/regression-ci-examples.md)
