---
title: "Drill: Simulate the CLT"
categories: [r, drill, coding-dojo]
tier-focus: script
filters:
  - webr
webr:
  packages:
    - ggplot2
    - tibble
---

# Drill: Simulate the Central Limit Theorem

## Drill Purpose

Build intuition for the sampling distribution of x̄ and the Central Limit Theorem by simulation. This underpins every CI and test statistic in the course.

## Trigger

Use this drill when you need to:

- Explain why we can use Z/t even when the population is not normal
- Demonstrate that SE = σ/√n shrinks with larger n
- Show that the sampling distribution of x̄ is approximately normal for large n

## Setup

```{r}
#| label: setup

pacman::p_load(pacman, tidyverse)
```

## Fixed Sequence

1. Define a non-normal population
2. Draw many samples of size n, compute x̄ for each
3. Plot the sampling distribution and overlay the theoretical normal
4. Compute SE and verify it matches sd(x̄ values)

## Worked Example

```{r}
#| label: clt-worked
#| fig-cap: "Sampling distribution of x̄: exponential population, n=30"

set.seed(42)

# Non-normal population: exponential with mean = 20
pop_rate <- 1/20
pop_mean <- 20
pop_sd   <- 20   # exponential: SD = mean

# Parameters
n_sim  <- 10000  # number of samples to draw
n_samp <- 30     # sample size

# Simulate
xbars <- replicate(n_sim, mean(rexp(n_samp, rate = pop_rate)))

# Theoretical SE
se_theory <- pop_sd / sqrt(n_samp)

cat("Theoretical SE = σ/√n =", round(se_theory, 4), "\n")
cat("Simulated  SE = sd(x̄) =", round(sd(xbars), 4), "\n")

# Plot
tibble(xbar = xbars) |>
  ggplot(aes(x = xbar)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = se_theory),
                color = "firebrick", linewidth = 1.2) +
  labs(title = sprintf("Sampling dist of x̄: Exponential pop, n = %d", n_samp),
       x = "Sample mean", y = "Density") +
  theme_bw()
```

## Practice Variation

Change `n_samp_new` below and click **Run Code** to see how the sampling distribution shape changes with sample size.

```{webr-r}
#| label: practice-interactive

library(ggplot2)
library(tibble)

pop_rate <- 1/20; pop_mean <- 20; pop_sd <- 20
n_sim <- 5000

n_samp_new <- 5    # try 5, 10, 30, 100

xbars_new <- replicate(n_sim, mean(rexp(n_samp_new, rate = pop_rate)))
se_new    <- pop_sd / sqrt(n_samp_new)

cat("SE = σ/√n =", round(se_new, 4), "\n")

tibble(xbar = xbars_new) |>
  ggplot(aes(x = xbar)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "coral", color = "white", alpha = 0.8) +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = se_new),
                color = "navy", linewidth = 1.2) +
  labs(title = sprintf("Sampling dist of x̄: n = %d, SE = %.2f", n_samp_new, se_new),
       x = "Sample mean", y = "Density") +
  theme_bw()
```

## Reflection Questions

1. At what sample size does the sampling distribution look approximately normal for an exponential population?
2. When you double n, by what factor does SE decrease?
3. Which part of the CLT result becomes the denominator in a Z-test or t-test?

## Related Wiki Pages

- [Distributions](../../wiki/concepts/distributions.md)
- [Standard Error](../../wiki/concepts/standard-error.md)
- [Confidence Intervals](../../wiki/concepts/confidence-intervals.md)
